---
title: "DME & VA Visit Visualization"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flat
date: "2025-04-10"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install.packages("kableExtra")

library(dplyr)
library(ggplot2)
library(ggridges)
library(kableExtra)
library(lubridate)
library(readr)
library(tidyr)

data_dir <- "/home/idies/workspace/home/SDoHinDRWestlundE/DME & VA/Visualization 03.31.2025"
data_csv <- "df_final_dme_va_03312025.csv"
raw_data_path <- file.path(data_dir, data_csv)

raw_data <- readr::read_csv(raw_data_path)


```

```{r functions, echo=FALSE, include = FALSE}
summarize_date <- function(data, date_var) {
  date_summary <- data |>
    summarise(
      min_date = min({{ date_var }}, na.rm = TRUE),
      max_date = max({{ date_var }}, na.rm = TRUE)
    )

  range_summary <- data |>
    filter(!is.na({{ date_var }})) |>
    group_by(patient_id) |>
    summarise(
      range_days = as.numeric(
        difftime(max({{ date_var }}), min({{ date_var }}), units = "days")
      ),
      .groups = "drop"
    ) |>
    summarise(
      min_range_months = min(range_days, na.rm = TRUE) / 30.44,
      pct_25_range_months = quantile(range_days, 0.25, na.rm = TRUE) / 30.44,
      mean_range_months = mean(range_days, na.rm = TRUE) / 30.44,
      median_range_months = median(range_days, na.rm = TRUE) / 30.44,
      pct_75_range_months = quantile(range_days, 0.75, na.rm = TRUE) / 30.44,
      max_range_months = max(range_days, na.rm = TRUE) / 30.44,
      sd_range_months = sd(range_days, na.rm = TRUE) / 30.44
    )

  bind_cols(date_summary, range_summary) |>
    mutate(across(where(is.numeric), ~ round(.x, 2))) |>
    mutate(across(everything(), as.character)) |>
    pivot_longer(everything(), names_to = "variable", values_to = "value") |> 
    knitr::kable()
}


plot_visit_range_violin <- function(data, date_var, var_desc) {
  data |>
    filter(!is.na({{ date_var }})) |>
    group_by(patient_id) |>
    summarise(
      visit_range_months = as.numeric(difftime(max({{ date_var }}), min({{ date_var }}), units = "days")) / 30.44
    ) |>
    ggplot(aes(x = "", y = visit_range_months)) +
    geom_violin(scale = "count") +
    geom_boxplot(width = 0.1) +
    labs(
      title = paste("Distribution of Visit Ranges per Patient"),
      subtitle = var_desc,
      x = NULL,
      y = "Visit Range (Months)"
    ) +
    theme_minimal()
}


plot_n_visits_ridge <- function(data, date_var, var_desc) {
  data |>
    filter(!is.na({{ date_var }})) |>
    group_by(patient_id) |>
    summarise(
      first_year = lubridate::year(min({{ date_var }})),
      n_visits = n_distinct(encounter_id)
    ) |>
    ggplot(aes(x = n_visits, y = factor(first_year))) +
    ggridges::geom_density_ridges(scale = 1.5, rel_min_height = 0.01) +
    labs(
      title = "Distribution of Number of Visits by First Visit Year",
      subtitle = var_desc,
      x = "Number of Visits",
      y = "First Visit Year"
    ) +
    theme_minimal()
}

plot_gap_histogram <- function(data, date_var, var_desc) {
  data |>
    filter(!is.na({{ date_var }})) |>
    arrange(patient_id, {{ date_var }}) |>
    group_by(patient_id) |>
    mutate(lag_date = lag({{ date_var }})) |>
    filter(!is.na(lag_date)) |>
    mutate(gap_months = as.numeric(difftime({{ date_var }}, lag_date, units = "days")) / 30.44) |>
    ggplot(aes(x = gap_months)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "white") +
    labs(
      title = "Gaps Between Visits",
      subtitle = var_desc,
      x = "Months Between Visits",
      y = "Frequency"
    ) +
    theme_minimal()
}

plot_monthly_volume <- function(data, date_var, var_desc) {
  data |>
    filter(!is.na({{ date_var }})) |>
    mutate(month = lubridate::floor_date({{ date_var }}, "month")) |>
    count(month) |>
    ggplot(aes(x = month, y = n)) +
    geom_col() +
    labs(
      title = "Monthly Encounter Volume",
      subtitle = var_desc,
      x = "Month",
      y = "Count"
    ) +
    theme_minimal()
}

plot_visit_timeline_sample <- function(data, date_var, var_desc, n_patients = 100) {
  sample_ids <- data |>
    distinct(patient_id) |>
    slice_sample(n = n_patients) |>
    pull()

  data |>
    filter(patient_id %in% sample_ids) |>
    ggplot(aes(x = {{ date_var }}, y = reorder(patient_id, {{ date_var }}))) +
    geom_point(size = 1) +
    labs(
      title = "Visit Timeline per Patient",
      subtitle = var_desc,
      x = "Visit Date",
      y = NULL
    ) +
    theme_minimal() +
    theme(axis.text.y = element_blank())
}

plot_visit_timeline_facet <- function(data, date_var, var_desc, n_patients_per_year = 50) {
  sampled_data <- data |>
    filter(!is.na({{ date_var }})) |>
    group_by(patient_id) |>
    summarise(first_year = lubridate::year(min({{ date_var }})), .groups = "drop") |>
    group_by(first_year) |>
    slice_sample(n = n_patients_per_year) |>
    inner_join(data, by = "patient_id")

  sampled_data |>
    ggplot(aes(x = {{ date_var }}, y = reorder(patient_id, {{ date_var }}))) +
    geom_point(size = 1) +
    labs(
      title = "Visit Timeline per Patient — Faceted by First Visit Year",
      subtitle = var_desc,
      x = "Visit Date",
      y = NULL
    ) +
    theme_minimal() +
    theme(axis.text.y = element_blank()) +
    facet_wrap(~ first_year, scales = "free_y")
}


```

# Data Prep

```{r}

data <- raw_data |>
  rename(
    patient_id = emrn,
    encounter_id = enc_id,
    cst_visit_date = visit_date,
    eye = eye,
    microscope_type = microscope,
    cst = central_subfield_thickness_num,
    va_method = method,
    va_logmar = BCVA,
    va_visit_date = cur_value_datetime,
    anti_vegf_received = code,
    anti_vegf_medication = med_name,
    anti_vegf_tx_date = proc_date
  ) |> mutate(
    patient_id = as.character(patient_id),
    encounter_id = as.character(encounter_id),
    cst_visit_date = as.Date(cst_visit_date),
    eye = factor(eye),
    microscope_type = as.factor(microscope_type),
    cst = as.numeric(cst),
    va_method = factor(va_method),
    va_logmar = as.integer(va_logmar),
    va_visit_date = as.Date(va_visit_date),
    anti_vegf_tx_date = as.Date(anti_vegf_tx_date),
    anti_vegf_received = as.logical(!is.na(anti_vegf_received)),
    anti_vegf_tx_date = as.Date(anti_vegf_tx_date)
  )

```

# Data File Stats

## Total N

```{r}

data |> 
  summarise(n_patients = n_distinct(patient_id)) |> 
  kable()

```

## Total Encounters

```{r}

data |>
  summarise(n_encounters = n_distinct(encounter_id)) |> 
  kable()

```

## Mean Encounters/Person

```{r}

data |> 
  group_by(patient_id) |> 
  summarise(enc_count = n_distinct(encounter_id)) |> 
  summarise(mean_encounters_per_person = mean(enc_count)) |> 
  kable()

```

# Central Subfield Thickness

## Visit Count

```{r}

data |> 
  filter(!is.na(cst_visit_date)) |> 
  summarise(n_with_cst = n()) |> 
  kable()

```

## Visit Date Summary

```{r}

data |> summarize_date(cst_visit_date)

```

## Visualization 

### Violin Plot of Visit Range

```{r}

data |> plot_visit_range_violin(cst_visit_date, "Visits With CST Measurement")

```

### Ridgeline Plot Of Number Of Visits

```{r}

data |> plot_n_visits_ridge(cst_visit_date, "Visits With CST Measurement")

```

### Gaps Between Visits

```{r}

data |> plot_gap_histogram(cst_visit_date, "Visits With CST Measurement")

```

### Monthly Encounter Volume

```{r}

data |> plot_monthly_volume(cst_visit_date, "Visits With CST Measurement")

```

### Timeline Per Patient

Random sample of 100 patients

```{r}

data |> plot_visit_timeline_sample(cst_visit_date, "Visits With CST Measurement")

```

### Visit Timeline Per Patient

50 sampled visits each

```{r, fig.height=20}

data |> plot_visit_timeline_facet(cst_visit_date, "Visits With CST Measurement")

```


# Visual Acuity

## Visit Count

```{r}

data |> 
  filter(!is.na(va_visit_date)) |> 
  summarise(n_with_va = n()) |> 
  kable()

```

## Visit Date Summary

```{r}

data |> summarize_date(va_visit_date)

```

## Visualization 

### Violin Plot of Visit Range

```{r}

data |> plot_visit_range_violin(va_visit_date, "Visits With Visual Acuity Measurement")

```

### Ridgeline Plot Of Number Of Visits

```{r}

data |> plot_n_visits_ridge(va_visit_date, "Visits With Visual Acuity Measurement")

```

### Gaps Between Visits

```{r}

data |> plot_gap_histogram(va_visit_date, "Visits With Visual Acuity Measurement")

```

### Monthly Encounter Volume

```{r}

data |> plot_monthly_volume(va_visit_date, "Visits With Visual Acuity Measurement")

```

### Timeline Per Patient

Random sample of 100 patients

```{r}

data |> plot_visit_timeline_sample(va_visit_date, "Visits With Visual Acuity Measurement")

```

### Visit Timeline Per Patient

50 sampled visits each

```{r, fig.height=20}

data |> plot_visit_timeline_facet(va_visit_date, "Visits With Visual Acuity Measurement")

```



# VEGF Procedure

## Visit Count

```{r}

data |> 
  filter(!is.na(anti_vegf_tx_date)) |> 
  summarise(n_with_vegf_procedure = n()) |> 
  kable()

```

## Visit Date Summary

```{r}

data |> summarize_date(anti_vegf_tx_date)

```

## Visualization 

### Violin Plot of Visit Range

```{r}

data |> plot_visit_range_violin(anti_vegf_tx_date, "Visits With Anti-VEGF Procedure")

```

### Ridgeline Plot Of Number Of Visits

```{r}

data |> plot_n_visits_ridge(anti_vegf_tx_date, "Visits With Anti-VEGF Procedure")

```

### Gaps Between Visits

```{r}

data |> plot_gap_histogram(anti_vegf_tx_date, "Visits With Anti-VEGF Procedure")

```

### Monthly Encounter Volume

```{r}

data |> plot_monthly_volume(anti_vegf_tx_date, "Visits With Anti-VEGF Procedure")

```

### Timeline Per Patient

Random sample of 100 patients

```{r}

data |> plot_visit_timeline_sample(anti_vegf_tx_date, "Visits With Anti-VEGF Procedure")

```

### Visit Timeline Per Patient

50 sampled visits each

```{r, fig.height=20}

data |> plot_visit_timeline_facet(anti_vegf_tx_date, "Visits With Anti-VEGF Procedure")

```